Robertson and Cao Project

### 2024-04-18 code/analysis-code/analysis1.qmd
Updated manuscript analysis
Examined problems with analysis in Rf and with LASSO. Added to manuscript.
I think that neither might be more suitable and we should try elastic net.

```
The residual plot shows clear collinearity that is not represented by the current model. To find the predictors with collinnearity we will use a correlation matrix. This is because the VIF analysis will not run when multicollinearity is too high.
```{r}
factor_data <- model.matrix(~ STATE + FACILITY + RACEETHK + AGE + EDUC1 + INS_BREAK_I + INCPOV1 + SEX + LANGUAGE + MOBIL_1 + INS_STAT2_I, data = mydata2)

@@

# Print variable pairs with correlation values of 1 or 0
print(variable_pairs)

```
This shows that many levels from each of the variables show multicollineaity. This might be addressed by using a different model for analysis.

>>>>

```
The residual plot shows clear collinearity that is not represented by the current model. To find the predictors with multicollinearity, we will use a correlation matrix. This is because the VIF analysis will not run when multicollinearity is too high.
```{r}
factor_data <- model.matrix(~ STATE + FACILITY + RACEETHK + AGE + EDUC1 + INS_BREAK_I + INCPOV1 + SEX + LANGUAGE + MOBIL_1 + INS_STAT2_I, data = mydata2)

@@

# Print variable pairs with correlation values of 1 or 0
print(variable_pairs)
```
This shows that many levels from each of the variables show multicollineaity. This might be addressed by using a different model for analysis.

### 2024-04-18 code/analysis-code/analysis2.5.qmd
Examined problems with analysis in Rf and with LASSO. Added to manuscript.
I think that neither might be more suitable and we should try elastic net.

## I used ChatGPT to help me define the tree grid for the parameters mtry and min_n
tree_grid <- grid_regular(mtry(range = c(1, 13)), 
                          min_n(range = c(1, 200)),
                          levels = 6)

@@

print(rf_vars)
```
The most important variable, according to this model, is state, followed by income poverty ratio, then facility, then facility. Despite the importance above, it should be noted that the multicollinearity may affect the importance of the variables selected. This means that a model that regularizes the parameters could be ideal for finding the most important predictors.
For this, we will try using the LASSO model

>>>>

## I used ChatGPT to help me define the tree grid for the parameters mtry and min_n
tree_grid <- grid_regular(mtry(range = c(1, 13)), 
                          min_n(range = c(1, 500)),
                          levels = 6)

@@

print(rf_vars)
```
The most important variable, according to this model, is state, followed by income poverty ratio, then facility, then facility. Despite the importance above, it should be noted that the multicollinearity may affect the importance of the variables selected. This means that a model that regularizes the parameters could be ideal for finding the most important predictors.
For this, we will try using the LASSO model.

### 2024-04-18 code/analysis-code/analysis3.qmd
Examined problems with analysis in Rf and with LASSO. Added to manuscript.
I think that neither might be more suitable and we should try elastic net.

# Prepare the recipe
lasso_prep <- lasso_rec %>%
  prep()
```
Now I will implement the function. This involves setting the specific LASSO regression engine, creating a workflow, and then extracting the fit of the model.
```{r}
# Specify logistic regression model
lasso_spec <- logistic_reg(penalty = 0.05, mixture = 1) %>%
  set_engine("glmnet")

@@

@@

@@

# LASSO Re-fit
Now, we will run the LASSO regression again, but with the tuned penalty parameter of 0.01. 
```{r}
# Specify logistic regression model
lasso_spec3 <- logistic_reg(penalty = 0.01, mixture = 1) %>%
  set_engine("glmnet")

# Fit the model using training data
lasso_fit3 <- lasso_wf %>%
  add_model(lasso_spec3) %>%
  fit(data = train_data)

# Extract and tidy model parameters
lasso_fit3 %>%
  extract_fit_parsnip() %>%
  tidy()

# Extract and tidy model parameters
lasso_coef <- lasso_fit3 %>%
  extract_fit_parsnip() %>%
  tidy()

@@

# Make predictions for ROC_AUC with the tuned model
predictions3 <- lasso_fit3 %>%
  augment(new_data = train_data)

@@

# Comparing Predictions
Now we will compare the two lasso models with the penalty parameters of 0.005 and 0.01 to determine which model performs best when presented with "new data". The new data is that split from the original data (the test data). 
```{r}
pen1_aug <- augment(lasso_fit2, test_data)
pen2_aug <- augment(lasso_fit3, test_data)
# Make find performance metrics with the tuned model
metrics <- metric_set(accuracy, f_meas) # create a set of metrics to test for classification
# Model tuned to parameter penalty = 0.005
penaltytune1 <- metrics(truth = P_UTDHPV, estimate = .pred_class, data = pen1_aug)
print(penaltytune1)
# Model tuned to parameter penalty = 0.01
penaltytune2 <- metrics(truth = P_UTDHPV, estimate = .pred_class, data = pen2_aug)
print(penaltytune2)
```
The performance metrics reveal a similar accuracy between both models, but a higher, thus better, F1 measure, from the LASSO model with the penalty of 0.005. This might indicate that, although the model tuned to the penalty 0.01 maximizes roc_auc, it may be at the cost of other performance metrics such as F1 (or precision and/or recall).
We will examine this further with confusion matrices.
```{r}
# Compare confusion matrices
# penalty 0.005
conf_mat(
  data = pen1_aug,
  truth = P_UTDHPV,
  estimate = .pred_class,
  dnn = c("Prediction", "Truth"),
  case_weights = NULL)
# penalty 0.01
conf_mat(
  data = pen2_aug,
  truth = P_UTDHPV,
  estimate = .pred_class,
  dnn = c("Prediction", "Truth"),
  case_weights = NULL)
```
The confusion matrix for neither of these models is ideal, as they both have a fairly high number of false positives, leading to a poor sensitivity. Because the model with a lower penalty parameter seems to perform better when presented with new data, I will look at the coefficients chosen by this model and determine which variables it deems most important to predicting HPV vaccine completion.
```{r}
# Finding importance of variables chosen
lasso_vars <- lasso_fit2 %>%
  extract_fit_parsnip() %>% # use parsnip to extract to fit
  vip::vip() #use vip for visualizing the most important variables

print(lasso_vars)
```
It seems that the most important factors for HPV vaccine completion are the state, facility and housing status. The states that have the highest predictive power of HPV vaccine completion are Rhode Island, Vermont, Mississippi, and North Dakota.

>>>>

# Prepare the recipe
lasso_prep <- lasso_rec %>%
  prep()

```
Now I will implement the function. This involves setting the specific LASSO regression engine, creating a workflow, and then extracting the fit of the model.
```{r}
# Specify logistic regression model
lasso_spec <- logistic_reg(penalty = 0.05, mixture = 1) %>%
  set_engine("glmnet")

@@

set.seed(123) # ste seed for reproducability

@@

# find the best penalty for ROC AUC
best_roc <- lasso_tuned %>%
  select_best(metric = "roc_auc")
best_roc

@@

# LASSO Re-fit
Now, we will run the LASSO regression again, but with the tuned penalty parameter of 0.001. 
```{r}
lasso_tune_spec <- logistic_reg(penalty = 0.001, mixture = 1) %>%
  set_engine("glmnet")

# Create workflow
lasso_tune_wf <- workflow() %>%
  add_recipe(lasso_rec)%>%
  add_model(lasso_tune_spec)

# Fit the model using training data
lasso_tune_fit <- lasso_tune_wf %>%
  fit(data = train_data)

# Extract and tidy model parameters
lasso_tune_coef <- lasso_tune_fit %>%
  extract_fit_parsnip() %>%
  tidy()

# Filter out predictors with non-zero coefficients
lasso_nonzero_coef <- lasso_tune_coef %>%
  filter(estimate != 0)

@@

# Make predictions for ROC_AUC with the tuned model
predictions3 <- lasso_tune_fit %>%
  augment(new_data = train_data)

@@

# Comparing Predictions
Now we will compare the two lasso models with the penalty parameters of 0.005 and 0.001 to determine which model performs best when presented with "new data". The new data is that split from the original data (the test data). 
```{r}
pen1_aug <- augment(lasso_fit2, test_data)
pen2_aug <- augment(lasso_tune_fit, test_data)
# Make find performance metrics with the tuned model
metrics <- metric_set(accuracy, f_meas) # create a set of metrics to test for classification
# Model tuned to parameter penalty = 0.005
penaltytune1 <- metrics(truth = P_UTDHPV, estimate = .pred_class, data = pen1_aug)
print(penaltytune1)
# Model tuned to parameter penalty = 0.01
penaltytune2 <- metrics(truth = P_UTDHPV, estimate = .pred_class, data = pen2_aug)
print(penaltytune2)
```
The performance metrics reveal a similar accuracy between both models, but a higher, thus better, F1 measure, from the LASSO model with the penalty of 0.001. 
```{r}
# Compare confusion matrices
# penalty 0.005
conf_mat(
  data = pen1_aug,
  truth = P_UTDHPV,
  estimate = .pred_class,
  dnn = c("Prediction", "Truth"),
  case_weights = NULL)
# penalty 0.001
conf_mat(
  data = pen2_aug,
  truth = P_UTDHPV,
  estimate = .pred_class,
  dnn = c("Prediction", "Truth"),
  case_weights = NULL)
```
The confusion matrix for neither of these models is ideal, as they both have a fairly high number of false positives, leading to a poor sensitivity. Because the model with a lower penalty parameter seems to perform better when presented with new data, I will look at the coefficients chosen by this model and determine which variables it deems most important to predicting HPV vaccine completion.
```{r}
# Finding importance of variables chosen
lasso_vars <- lasso_tune_fit %>%
  extract_fit_parsnip() %>% # use parsnip to extract to fit
  vip::vip() #use vip for visualizing the most important variables

print(lasso_vars)
```
It seems that the most important factors for HPV vaccine completion are the state, facility and housing status. The states that have the highest predictive power of HPV vaccine completion are Rhode Island, Vermont, Mississippi, and North Dakota.
```{r}
# Extract and tidy model parameters
lasso_tune_coef <- lasso_tune_fit %>%
  extract_fit_parsnip() %>%
  tidy()

# Filter out predictors with non-zero coefficients
lasso_tune_coef <- lasso_tune_coef %>%
  filter(estimate != 0)
lasso_tune_coef
```

### 2024-04-18 products/manuscript/Manuscript.qmd
Examined problems with analysis in Rf and with LASSO. Added to manuscript.
I think that neither might be more suitable and we should try elastic net.

# Results

## Exploratory Data Analysis

Exploratory analysis is used to determine the distribution of our predictors, including the socioeconomic variables that we chose for analysis.
Nearly all of the demographic factors have a similar distribution between those who have completed their HPV vaccination regime versus those who have not completed it.

```{r}
#| label: Table1
#| tbl-cap: "Table 1.Socioeconomic characteristics of U.S. teenagers who have completed or are up-to-date with HPV vaccination series."
#| echo: FALSE
Table1 <- readRDS(here("results/tables/table1.rds"))
kable(Table1)
```

The distribution of the numerical income to poverty ratio can be shown below.
It displays a value from 1-3.
A ratio displaying less than 1 depicts an income less than the poverty level.Any income ratio \>1 indicates an income greater than the poverty level.
Most of the observations were gathered from households with income ratios of 3, as shown by the skewed plot below.

@@

Next, the distribution of surveys given per state was analyzed.
The state variable shows a more uniformed distribution, suggesting that the observations were gathered fairly across the US states and territory.
Although it is noted a few territories provides 0 observations.
![Figure 2. State distribution](./images/state.distribution.png)

The family income variable provides a more normal distribution with the majority of the observation coming from \$75000+ income level.
![Figure 3. The family income distribution](./images/family.income.distribution.png)

The race and ethnicity variables show a distribution that is similar to the U.S. census data, which concludes that the racial composition of the U.S. is 58.9% white, 13.6% African American, and 19.1% Hispanic or Latino.

![Figure 4.Racial distribution](./images/race.distribution.png)

![Figure 5.Ethnicity distribution](./images/ethnicity.distribution.png)

The Insurance status variable shows that the majority of the observations came from households with private insurance only.
This is followed by any medicaid and then other insurance.
This is mirrored in the facility variable, where a majority of its observations show private facilities being the location.
![Figure 6. Insurance status distribution](./images/insurance.status.distribution.png) ![Figure 7. Distribution of the medical facility where the survey was administerred](./images/facility.distribution.png)

The vaccination variable indicates nearly 80% of the observations had an Up-to-date status for the 1+ shot HPV vaccination.
We will follow by examining the percentage of up-to-date vaccination status by state.

![Table 2. Vaccine status distribution](./images/state-vaccine-completion.png)

We examine the facility that submitted the survey, which is stratified by the vaccine completion status of the survey subject.

![Figure 8.Facility and vaccine completion distribution](./images/facility-vaccination.png)

We examine the percentage of vaccination completion by race and ethnicity.
Because there are different counts of each race/ethnicity for each factor level, the percentage of those with up-to-date vaccine status out of each race/ethnicity level was determined.
This adjusts for the different counts in each racial category.
The Hispanic ethnicity was found to have the highest vaccine completion while non-hispanic white has the lowest.

![Figure 9. Distribution of vaccine completion stratified by race and ethnicity](./images/raceeth-income-stratified.png)

## Model and Plot Fitting

A boxplot is created with income-poverty ratio to vaccination status.
The box plot indicates little difference between the two status.
A statistical model is fitted to the same variables.
A p-value of 0.1358 suggests there is little significance between the income-poverty variable and the vaccination status.

![Figure 10.](./images/income-vaccination.png)

A boxplot is created with income-poverty ratio to insurance status.
The box plot indicates there are distinct difference between the two status.
A statistical model is fitted to the same variables.
The p-value of \< 2.26e-16 suggests there is statistical significance between the income-poverty variable and the insurance status.

@@

Income-poverty ratio was fit to vaccine status and insurance status in a multi-linear model.

A box-plot was created for income-poverty level versus race/ethnicity and this was stratified by vaccine completion (UTD or NOT UTD).
A lower income-poverty ratio means the group is closer to the poverty line.
For the non-Hispanic black and Hispanic racial and ethnic groups, vaccine completion corresponded with a lower mean income-poverty ratio.
For mixed ethnic groups, vaccine completion corresponded with a higher mean income-poverty ratio.

![Figure 12. Income-poverty level versus ethnic group](./images/raceeth-income-stratified.png)

@@

We begin with various socioeconomic predictors that interest us.
Some of these predictors are a subset of the other, or may be used to describe an overall category.

>>>>

# Results
## Exploratory Data Analysis
Exploratory analysis is used to determine the distribution of our predictors, including the socioeconomic variables that we chose for analysis. Nearly all of the demographic factors have a similar distribution between those who have completed their HPV vaccination regime versus those who have not completed it.

```{r}
#| label: Table1
#| tbl-cap: "Table 1.Socioeconomic characteristics of U.S. teenagers who have completed or are up-to-date with HPV vaccination series."
#| echo: FALSE
Table1 <- readRDS(here("results/tables/table1.rds"))
kable(Table1)
```
Income-poverty-ratio is a numeric value that indicates how close to poverty a family is living. It is calculated by income over the poverty income threshold and ranges from values 1 (closest to poverty) and 3 (furthest from poverty). The distribution of the income poverty ratio indicates a strong left skew. The highest frequency of observations was gathered from households with an income poverty ratio, meaning most survey participants live well above the poverty threshold. The family income is right skewed with the majority of the observations coming from \$75000+ income level. This indicates that a high frequency of participants are above the poverty income threshold, but also in a modest income class.

@@

![Figure 2. The family income distribution](./images/family.income.distribution.png)
The state variable shows a more uniform distribution, suggesting that the observations were gathered fairly across the US states and territories. States and territories with a significantly larger number of observations include Texas, West Virginia, Pennsyvania, Puerto Rico, New York, and Illinois. Although it is noted a few territories provide 0 observations. The race and ethnicity variables show a distribution that is similar to the U.S. census data. The census data states that the racial composition of the U.S. is 58.9% white, 13.6% African American, and 19.1% Hispanic or Latino, while our survey population is composed of 58.% white, 9.15% African American, and 19.9% Hispanic or Latino participants.
The majority of observations came from households with private insurance only The Insurance status variable shows that the majority of the observations came from households with private insurance only (58.6%), followed by Medicaid (31.6%), other insurance (7.7%), and uninsured (2.2%). This is mirrored in the facility variable, as survey location, where a majority of its observations show private facilities (34.4%), followed by mixed facilities (25.8%), hospitals (11.3%), and all public facilities (9.9%). 15.2% of the facility types were unknown. Upon stratification of facility type by HPV vaccination completion status,  it was found that the majority of participants who were up-to-date were surveyed at private or mixed facilities. They likely received their HPV vaccination there as well.
![Figure 3.Facility and vaccine completion distribution](./images/facility-vaccination.png)
A total of 79% of participants had up-to-date status for receiving 1 or more shots of HPV vaccination. U.S. states and territories had an HPV vaccination completion rate ranging from 69% to 95%, fluctuating around ~80%. There are different counts of each race/ethnicity for each factor level, so the percentage of those with up-to-date vaccine status out of each race/ethnicity level was adjusted. The Hispanic ethnicity was found to have the highest vaccine completion while the non-Hispanic white ethnicity had the lowest. It was also found that for Hispanic and Non-Hispanic black ethnicities, the income poverty ratio was higher for those who were not up-to-date in HPV vaccination status, compared to those who were. The opposite was true for mixed ethnic groups, where vaccine completion corresponded with a higher mean income-poverty ratio. When examining the overall HPV vaccination status stratified by income poverty ratio, the income poverty ratio did not differ between those who are up to date and those who are not. This may be accounted for by other trends such as that shown with ethnicity, above.
![Figure 9. Distribution of income poverty ratio stratified by race and HPV vaccination completion status](./images/raceeth-income-stratified.png)
	Subsequently, income poverty ratio and insurance status were examined, and stratified by HPV vaccination status. It is apparent that those with private and other insurance status had the highest income-poverty ratio, while those with Medicaid or uninsured were closer to the poverty income threshold. For those with Medicaid, those who were up-to-date in HPV vaccinations had a slightly lower income-poverty ratio than those who were not up to date.

@@

@@

We begin with various socioeconomic predictors that interest us.
Some of these predictors are a subset of the other, or may be used to describe an overall category.
